Reagents: expressing and composing fine-grained concurrency

### 2011-10-18 paper/paper.tex
paper
work on Reagents explanation

\section{Background}

@@

@@

In the last section, we saw a range of strategies for communication
and synchronizion at a fine grain:
\begin{itemize}
\item using CAS to make direct, atomic updates;
\item using exponential backoff to ease contention; 
\item using side-channel communication to increase parallelism; 
\item using disjoint spinwaiting to eliminate useless cache traffic; 
\item using dual data structures to incorporate blocking.
\end{itemize}
Reagents incorporate these disparate techniques into a uniform framework for
writing concurrent algorithms.  Some of the techniques are built into the
framework, while others are expressible as reusable abstractions.

more easily and concisely
% extensibility and composability

\subsection{The main idea}

Reagents are a new instance of an old idea: computations can be usefully
represented as data.

% no reason not to present this as an ADT

@@

chemistry metaphor

@@

\subsection{Expressiveness by example}

\begin{lstlisting}
class TreiberStack[A] {
  private val head = new Ref[List[A]](Nil)
  val push: Reagent[A, Unit] = head.upd {
    (xs, x) => (x:xs, ())
  }
  val tryPop: Reagent[Unit, Option[A]] = head.upd {
    case (x::xs, ()) => (xs, Some(x))
    case (Nil,   ()) => (Nil, None)
  }
}
\end{lstlisting}

@@

\begin{lstlisting}
class MSQueue[A] {
  private case class Node(data: A, next: Ref[Node])
  private val head = Ref(Node(null))
  private val tail = Ref(head.read!())

  @tailrec private def search(x:A): Reagent[Unit,Unit] = 
    tail.read ! () match {
      case Node(_, r@Ref(null)) => 
        r.cas(null, Node(x, new Ref(null)))
      case ov@Node(_, Ref(nv)) => 
        tail.cas(ov,nv) !? (); search(x)
    }
  val enq: Reagent[A, Unit] = computed search
  val tryDeq: Reagent[Unit, Option[A]] = head.upd {
    case Node(_, Ref(n@Node(x, _))) => (n, Some(x))
    case emp => (emp, None)
  }
}
\end{lstlisting}

@@

@@

@@

CML, Transactional events, join calculus

>>>>

\section{Background}
\label{sec:background}

@@

% OPTIMISTIC CONCURRENCY

@@

Reagents are a new instance of an old idea: representing computations as data.
The computations being represented are fine-grained concurrent operations, so
a value of type \reagent{A}{B} represents a function from \lstinline{A} to
\lstinline{B} that internally interacts with a concurrent data structure
through mutation, synchronization, or both.

Unlike $\lambda$-abstractions, which encompass arbitrary code and treat that
code opaquely, reagents are built up from a limited set of \emph{combinators}
that are known to the implementation.  Existing reagents---for example, those
built by a concurrency expert---can be composed or tailored by library users
by applying additional combinators, without knowledge of the internal
implementation.  At any point a user can invoke a reagent, either as a
\emph{reactant} or a \emph{catalyst}---a distinction explained in
\secref{easy-examples}.

Reagents are patterned after Concurrent ML~\cite{?} and Haskell's STM
monad~\cite{?}, which represent (respectively) synchronous and atomic
operations as first-class values; we discuss the relationship in
\secref{related}.
%
Our contribution is giving a set of combinators appropriate for expressing and
composing fine-grained concurrent algorithms (\secref{easy-examples}) with a
clear semantics (\secref{semantics}), ... %TODO

%more easily and concisely
% extensibility and composability

@@

\subsection{Basic reagents: updates, messages, and choice}
\label{sec:easy-examples}

In \secref{background}, we saw a range of strategies for communication and
synchronizion at a fine grain:
\begin{itemize}
\item using CAS to make direct, optimistic, and atomic updates;
\item using exponential backoff to ease contention; 
\item using side-channel communication to increase parallelism; 
\item using disjoint spinwaiting to eliminate useless cache traffic; 
\item using dual data structures to incorporate blocking.
\end{itemize}
Reagents incorporate these disparate techniques into a uniform set of
combinators for writing concurrent algorithms.  Some of the techniques are
built into the combinators themselves, while others are expressible as
reusable abstractions.

\begin{lstlisting}
class TreiberStack[A] {
  private val head = new Ref[List[A]](Nil)
  val push: Reagent[A, Unit] = upd(head) {
    case (xs, x) => (x:xs, ())
  }
  val tryPop: Reagent[Unit, Option[A]] = upd(head) {
    case (x::xs, ()) => (xs, Some(x))
    case (Nil,   ()) => (Nil, None)
  }
}
\end{lstlisting}

The first two techniques are built into the combinators.  We introduce a new
type \lstinline{Ref[A]} for atomically updatable references holding data of
type \lstinline{A}.  Unlike Scala's (and Java's) \lstinline{AtomicReference},
\lstinline{Ref} values can only be accessed through the following reagent
combinators:
\begin{lstlisting}[mathescape=true]
read: Ref[A] => Reagent[Unit, A]
cas:  Ref[A] * A * A => Reagent[Unit, Unit]
upd:  Ref[A] => PartialFunction[(A,B), (A,C)] 
             => Reagent[B, C]
\end{lstlisting}
By restricting access to the combinators, ... %TODO

% Freedom to change kcas implementations

The \lstinline{read} combinator has a straightforward signature: if
\lstinline{r} has type \lstinline{Ref[A]}, then \lstinline{read(r)} is a
\emph{value} of type \lstinline{Reagent[Unit, A]} which, when invoked, takes a
unit\footnote{\lstinline{Unit} in Scala is akin to \lstinline{void}: it is a
  type with a single member, written \lstinline{()}.} argument and yields an
\lstinline{A} result.

The \lstinline{cas} combinator takes two additional \lstinline{A} arguments,
giving the expected and updated values, respectively.  Unlike its counterpart
for \lstinline{AtomicReference}, a \lstinline{cas} reagent does \emph{not}
yield a boolean result.  Instead, the reagent library builds in handling of
failed CASes, including an optimistic retry loop and exponential backoff.  The
failure mechanics depend on which other combinators are involved, so we delay
a complete discussion until \secref{failure}.

The \lstinline{upd} combinator

%% THIS SEEMS MORE CONFUSING THAN HELPFUL
%% For now, it suffices to know that
%% invoking a \lstinline{cas} will complete and return \lstinline{()} only when
%% the CAS has successfully completed.

%% \begin{lstlisting}[mathescape=true]
%% def upd[A,B,C](r: Ref[A], f: (A,B) $\rightharpoonup$ (A,C)): Reagent[B,C] 
%%   = computed { 
%%       (b: B) =>       
%% \end{lstlisting}

chemistry metaphor.  note that catalysts increase parallelism and remove need
for devoted thread

@@

key point: traditional lock-free algorithms that use a single cas do not pay
any penalty for composability/kcas potential

show how, for counter, tryDec and dec are inter-expressible

% TODO: fix types for cases below
\begin{lstlisting}
class Counter {
  private val state = new Ref[Int](0)
  val get = upd(state) { case i => (i, i) }
  val inc = upd(state) { case i => (i+1, i) }
  val dec = upd(state) { case i if (i > 0) => (i-1, i) }
  val tryDec = upd(state) {
     case i if (i == 0) => (0,   None)
     case i              => (i-1, Some(i))
  }
}
\end{lstlisting}

% nice catalysts: zip and unzip

\subsection{Failure is an \textsf{Option}}
\label{sec:failure}

blocking failure versus retryable failure

\begin{lstlisting}
  val tryDec = 
\end{lstlisting}

\subsection{Advanced reagents: CAS, computed, and composed}
\label{sec:hard-examples}

@@

\begin{lstlisting}
class MSQueue[A] {
  private case class Node(data: A, next: Ref[Node])
  private val head = new Ref(Node(null))
  private val tail = new Ref(head.read!())

  @tailrec private def search(x:A): Reagent[Unit,Unit] = 
    read(tail) ! () match {
      case Node(_, r@Ref(null)) => 
        cas(r, null, Node(x, new Ref(null)))
      case ov@Node(_, Ref(nv)) => 
        cas(tail, ov, nv) !? (); 
        search(x)
    }
  val enq: Reagent[A, Unit] = computed search
  val tryDeq: Reagent[Unit, Option[A]] = upd(head) {
    case Node(_, Ref(n@Node(x, _))) => (n, Some(x))
    case emp => (emp, None)
  }
}
\end{lstlisting}

post-commit action for catching up tail?

@@

\begin{lstlisting}
def !(a: A): B = tryReact(a, Reaction.inert, null) match {
  case (_: BacktrackCommand) => {
    val backoff = new Backoff
    val maySync = this.maySync // cache
    @tailrec def retryLoop(shouldBlock: Boolean): B = {
      val wait = maySync || shouldBlock
      val waiter = if (wait) new Waiter[B](shouldBlock) 
                   else null

      tryReact(a, Reaction.inert, waiter) match {
        case (bc: BacktrackCommand) if wait => {
          bc.bottom(waiter, backoff, snoop(a))
          waiter.tryAbort      // rescind waiter,
          waiter.poll match { // but check if completed
            case Some(ans) => ans.asInstanceOf[B] 
            case None       => retryLoop(bc.isBlock)
          }
        }
        case Retry => backoff.once; retryLoop(false)
        case Block => retryLoop(true)
        case ans   => ans.asInstanceOf[B]
      }
    }
    retryLoop(false)
  }
  case ans => ans.asInstanceOf[B]
}
\end{lstlisting}

@@

\begin{lstlisting}
  def tryReact(u: Unit, rx: Reaction, offer: Offer[B]): Any = 
    if (rx.canCASImmediate(k, offer)) {
      if (casI(expect, update))
        k.tryReact((), rx, offer)
      else Retry
    } else k.tryReact((), rx.withCAS(Ref.this, expect, update), offer)
\end{lstlisting}

@@

CML, Transactional events, async CML, join calculus
